---
title: "Exploring Sodol's indices and randomness"
shorttitle: "Exploring Sodol's indices and randomness"
author:
- name: "Frédéric Bertrand"
  affiliation:
  - Cedric, Cnam, Paris
  email: frederic.bertrand@lecnam.net
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Exploring Sodol's indices and randomness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/sobol-randomness-",
  fig.width = 7,
  fig.height = 5,
  dpi = 150,
  message = FALSE,
  warning = FALSE
)

LOCAL <- identical(Sys.getenv("LOCAL"), "TRUE")

library(Sobol4R)
library(sensitivity)
library(ggplot2)
set.seed(4669)
```


# Overview
This vignette depictsis a first use of the Sobol4R package and the Sobol indices for models with both parameters for fixed and random effects.

If you are a Linux/Unix or a Macos user, you can install a version of Sobol4R with support for `doMC` from [github](https://github.com) with:

```{r, eval = FALSE}
devtools::install_github("fbertran/Sobol4R", ref = "doMC")
```

Since the simulation time is quite large, we shipped the results with the package as an example dataset. We provide the code to rerun the simulations at the end of the vignette.

### Context and non random case

#### Test case : the non-monotonic Sobol g-function

The method of sobol requires 2 samples (there are 2 variables, all following the uniform distribution on $[0,1$])

```{r}
library(Sobol4R)
set.seed(4669)
n <- 50000
#X1_1 <- data.frame(matrix(runif(8 * n), nrow = n))
data(X1_1, package="Sobol4R")
summary(X1_1)
#X2_1 <- data.frame(matrix(runif(8 * n), nrow = n))
data(X2_1, package="Sobol4R")
summary(X2_1)
```









```{r}
library(sensitivity)
if(requireNamespace("sensitivity", quietly = TRUE)){
sensitivity::sobol.fun
}
```

```{r}
library(sensitivity)
if(requireNamespace("sensitivity", quietly = TRUE)){
gensol1 <- sensitivity::sobol(model = NULL, X1 = X1_1, X2 = X2_1, order = 2, nboot = 100)
#Y1 = sensitivity::sobol.fun(gensol1$X)
data(Y1, package="Sobol4R")
}
```


```{r}
if(requireNamespace("sensitivity", quietly = TRUE)){
x1 <- sensitivity::tell(gensol1, Y1)
print(x1)
ggplot2::ggplot(x1)
rm(gensol1,X1_1,X2_1)
}
```



